---
title: "Correlations"
author: "Nadine Bestard"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## set-up
```{r}
library(DropletUtils)
library(here)
library(scuttle)
library(VennDiagram)
library(ggplot2)
library(dplyr)

 save_path <- here("outs/correlations")
 dir.create(save_path)
 
 colour_blue_yellow <- ggsci::pal_jco("default")(2)
```
# Load datasets
```{r}
#thalamus
sample_names_th_illumina <-dir(here("outs/cellranger/downsampled"), pattern = "^[T]") # extract TH and ZB
sample_names_th_mgi <-dir(here("outs/cellranger/downsampled"), pattern = "^MGI_[T]") # extract TH and ZB
# zb
sample_names_zb_illumina <-dir(here("outs/cellranger/downsampled"), pattern = "^[Z]") # extract  ZB
sample_names_zb_mgi <-dir(here("outs/cellranger/downsampled"), pattern = "^MGI_[Z]") # extract  ZB
#sampels
samples_th_illumina <- here(paste0("outs/cellranger/downsampled/", sample_names_th_illumina, "/outs/filtered_feature_bc_matrix/"))
samples_th_mgi <- here(paste0("outs/cellranger/downsampled/", sample_names_th_mgi, "/outs/filtered_feature_bc_matrix/"))
samples_zb_illumina <- here(paste0("outs/cellranger/downsampled/", sample_names_zb_illumina, "/outs/filtered_feature_bc_matrix/"))
samples_zb_mgi <- here(paste0("outs/cellranger/downsampled/", sample_names_zb_mgi, "/outs/filtered_feature_bc_matrix/"))

#into sce
sce_th_mgi <- read10xCounts(samples_th_mgi, sample.names = sample_names_th_mgi, col.names = TRUE)
sce_th_illumina <- read10xCounts(samples_th_illumina, sample.names = sample_names_th_illumina, col.names = TRUE)

#zb
sce_zb_mgi <- read10xCounts(samples_zb_mgi, sample.names = sample_names_zb_mgi, col.names = TRUE)
sce_zb_illumina <- read10xCounts(samples_zb_illumina, sample.names = sample_names_zb_illumina, col.names = TRUE)

```

# Zebra Fish 
## Add metrics
```{r}
# add quality info
sce_zb_mgi <- addPerCellQCMetrics(sce_zb_mgi)
sce_zb_mgi <- addPerFeatureQCMetrics(sce_zb_mgi)
sce_zb_illumina <- addPerCellQCMetrics(sce_zb_illumina)
sce_zb_illumina <- addPerFeatureQCMetrics(sce_zb_illumina)

# Sort duplicates in barcodes
sce_zb_illumina$Barcode_uniq <- colnames(sce_zb_illumina)
sce_zb_mgi$Barcode_uniq <- colnames(sce_zb_mgi)


```

## Compare

### num cells
```{r}
dim(sce_zb_illumina)[2]
dim(sce_zb_mgi)[2]

sum(colnames(sce_zb_illumina) %in% colnames(sce_zb_mgi))
sum(colnames(sce_zb_mgi) %in% colnames(sce_zb_illumina)) # just checking no dups
sum(sce_zb_illumina$Barcode_uniq %in% sce_zb_mgi$Barcode_uniq)
```


```{r, echo=FALSE}
venn.diagram(list(colnames(sce_zb_mgi), colnames(sce_zb_illumina)), 
             
             #aes
             col = colour_blue_yellow,
             fill = c(alpha(colour_blue_yellow[1]), alpha(colour_blue_yellow[2])),
             #save
             height = 4800,
             width = 7000,
             resolution = 300,
             filename = here(save_path, "vennZBilluminavsMGI_cells.png"),
             disable.logging=TRUE,
             #text nums
             cex=4,
             fontfamily="sans",
             #text categories
             category.names = c("MGI" , "Illumina"),
             cat.cex = 4,
             cat.fontfamily="sans",
             cat.default.pos="outer",
             cat.dist = c(0.055, 0.055),
             #title
             main="Cells",
             main.fontfamily = "sans",
             main.cex=5
             )
```


```{r}
# for some reason now I have 6 dups. I checked and they are in both, so must be dups. # colnames are unique
compare_cells <-  merge(colData(sce_zb_illumina), colData(sce_zb_mgi), by = "Barcode_uniq", suffixes = c("_illumina","_mgi"))
dim(compare_cells) 
```

## Correlation cells
```{r}
#plot(compare_cells$sum_illumina, compare_cells$sum_mgi)
ggplot(as.data.frame(compare_cells), aes(x=sum_illumina, y = sum_mgi)) + 
  geom_point()

lm <- lm(compare_cells$detected_illumina ~ compare_cells$detected_mgi)
r2 <- summary(lm)$r.squared
ggplot(as.data.frame(compare_cells), aes(x=detected_illumina, y = detected_mgi)) + 
 # geom_smooth(method = "lm", colour="grey") +
  geom_point(shape="o") +
  ggtitle("Correlation Cells", 
          subtitle = paste("Num Cells =", dim(compare_cells)[1])) +
  labs(x = "Illumina", y="MGI",
       caption = "Number of detected genes for each cell") +
 # geom_text(x = 3000, y = 10000, label = expression(r2))#(expression("R"^2)))#, round(r2,4))))
geom_text(x = 3000, y = 10000, label = expression(paste("R"^2, "= 0.9995"))) + # I really tried to use the variable, but it does not work with the R2 expression
theme_classic()

```

## number of genes
```{r}
# subset to only have genes expressed
sce_zb_illumina <- sce_zb_illumina[rowSums(assay(sce_zb_illumina))>0,]
sce_zb_mgi <- sce_zb_mgi[rowSums(assay(sce_zb_mgi))>0,]

dim(sce_zb_illumina)[1]
dim(sce_zb_mgi)[1]

compare_genes <- merge(rowData(sce_zb_illumina), rowData(sce_zb_mgi), by="ID", suffixes = c("_illumina", "_mgi"))
```


```{r venngenesZB, echo=FALSE}
venn.diagram(list(rownames(sce_zb_mgi), rownames(sce_zb_illumina)), 
             
             #aes
             col = colour_blue_yellow,
             fill = c(alpha(colour_blue_yellow[1]), alpha(colour_blue_yellow[2])),
             #save
             height = 4800,
             width = 7000,
             resolution = 300,
             filename = here(save_path, "vennZBilluminavsMGI_genes.png"),
             disable.logging=TRUE,
             #text nums
             cex=4,
             fontfamily="sans",
             #text categories
             category.names = c("MGI" , "Illumina"),
             cat.cex = 4,
             cat.fontfamily="sans",
             cat.default.pos="outer",
             cat.dist = c(0.055, 0.055),
             #title
             main="Genes",
             main.fontfamily = "sans",
             main.cex=5
             )
```

## correlation genes
```{r}
lm <- lm(compare_genes$mean_illumina ~ compare_genes$mean_mgi)
r2 <- summary(lm)$r.squared
r2
ggplot(as.data.frame(compare_genes), aes(x=mean_illumina, y=mean_mgi )) +
  geom_point(shape="o") +
  ggtitle("Gene correlation", 
          subtitle = paste("Num Genes =", dim(compare_genes)[1])) +
  labs(x= "Illumina", y= "MGI",
       caption = "Average expression of each gene across all cells") +
  
 # geom_smooth(method = "lm", colour = "black")
  geom_text(x = 25, y = 75, label = expression(paste("R"^2, "= 0.9969"))) +#, round(r2,4)))) # I really tried to use the variable, not possible
    theme_classic()
  
```

# Mouse 
## Add metrics
```{r}
# add quality info
sce_th_mgi <- addPerCellQCMetrics(sce_th_mgi)
sce_th_mgi <- addPerFeatureQCMetrics(sce_th_mgi)
sce_th_illumina <- addPerCellQCMetrics(sce_th_illumina)
sce_th_illumina <- addPerFeatureQCMetrics(sce_th_illumina)

sce_th_illumina$Barcode_uniq <- colnames(sce_th_illumina)
sce_th_mgi$Barcode_uniq <- colnames(sce_th_mgi)
```

## Compare

### num cells
```{r}
dim(sce_th_illumina)[2]
dim(sce_th_mgi)[2]

sum(colnames(sce_th_illumina) %in% colnames(sce_th_mgi))

#save the cells that are different from mgi and illumina
only_illumina <- colnames(sce_th_illumina)[!(colnames(sce_th_illumina) %in% colnames(sce_th_mgi))]
only_mgi <- colnames(sce_th_mgi)[!(colnames(sce_th_mgi) %in% colnames(sce_th_illumina))]

saveRDS(only_mgi, here(save_path, "cells_only_in_mgi.RDS"))
saveRDS(only_illumina, here(save_path, "cells_only_in_illumina.RDS"))
```

I saved the cells that are different between the two technologies to track them downstream

```{r venncellsTH}
venn.diagram(list(colnames(sce_th_illumina), colnames(sce_th_mgi)), 
             
             #aes
             col = colour_blue_yellow,
             fill = c(alpha(colour_blue_yellow[1]), alpha(colour_blue_yellow[2])),
             #save
             height = 4800,
             width = 7000,
             resolution = 300,
             filename = here(save_path, "vennTHilluminavsMGI_cells.png"),
             disable.logging=TRUE,
             #text nums
             cex=4,
             fontfamily="sans",
             #text categories
             category.names = c("MGI" , "Illumina"),
             cat.cex = 4,
             cat.fontfamily="sans",
             cat.default.pos="outer",
             cat.dist = c(0.055, 0.055),
             #title
             main="Cells",
             main.fontfamily = "sans",
             main.cex=5
             )
```

## Compare


```{r}

compare_cells <-  merge(colData(sce_th_illumina), colData(sce_th_mgi), by = "Barcode_uniq", suffixes = c("_illumina","_mgi"))
dim(compare_cells) # for some reason now I have 6 dups. I checked and they are in both, so must be dups. 
```
## Correlation cells
```{r}
# TODO improve this plot, and add it in the ZB
lm <- lm(compare_cells$sum_illumina ~ compare_cells$sum_mgi)
r2_sum <- summary(lm)$r.squared
r2_sum
ggplot(as.data.frame(compare_cells), aes(x=sum_illumina, y = sum_mgi)) + 
   geom_point(shape="o") +
  ggtitle("Correlation Cells", 
          subtitle = paste("Num Cells =", dim(compare_cells)[1])) +
  labs(x = "Illumina", y="MGI",
       caption = "Number of UMI counts for each cell") +
 geom_text(x = 20000, y = 50000, label = expression(paste("R"^2, "= 0.9961"))) + 
  theme_classic()

lm <- lm(compare_cells$detected_illumina ~ compare_cells$detected_mgi)
r2 <- summary(lm)$r.squared
r2

as.data.frame(compare_cells) %>% 
ggplot(aes(x=detected_illumina, y = detected_mgi)) + 
 # geom_smooth(method = "lm", colour="grey") +
  geom_point(shape="o") +
  ggtitle("Correlation Cells", 
          subtitle = paste("Num Cells =", dim(compare_cells)[1])) +
  labs(x = "Illumina", y="MGI",
       caption = "Number of detected genes for each cell") +
 geom_text(x = 2500, y = 7500, label = expression(paste("R"^2, "= 0.9964"))) + #, round(r2,4)))) # I really tried to use the variable, but it does not work with the R2 expression
  theme_classic()

# Same but marking outliers
as.data.frame(compare_cells) %>% 
  mutate(ratio = detected_illumina / detected_mgi) %>% 
  mutate(outlier = ifelse(ratio < quantile(ratio, 0.025) | ratio > quantile(ratio, 0.975), "TRUE", "FALSE")) %>% 
ggplot(aes(x=detected_illumina, y = detected_mgi, colour = outlier)) + 
 # geom_smooth(method = "lm", colour="grey") +
  geom_point(shape="o") +
   ggtitle(paste("Cells correlation")) + 
  labs(x = "Illumina", y="MGI",
       caption = "Number of detected genes for each cell") +
 geom_text(x = 2500, y = 7500, label = expression(paste("R"^2, "= 0.9964"))) + #, round(r2,4)))) # I really tried to use the variable, but it does not work with the R2 expression
  theme_classic()

 
 #save to file the outliers
 outlier_barcodes <- as.data.frame(compare_cells) %>% 
   mutate(ratio = detected_illumina / detected_mgi) %>% 
  mutate(outlier = ifelse(ratio < quantile(ratio, 0.025) | ratio > quantile(ratio, 0.975), "TRUE", "FALSE")) %>% 
   pull(Barcode_uniq)
 saveRDS(outlier_barcodes, here("outs/correlations/outlier_cells.RDS"))
```

The outlier cells are saved

## number of genes
```{r}
# subset to only have genes expressed
sce_th_illumina <- sce_th_illumina[rowSums(assay(sce_th_illumina))>0,]
sce_th_mgi <- sce_th_mgi[rowSums(assay(sce_th_mgi))>0,]

dim(sce_th_illumina)[1]
dim(sce_th_mgi)[1]

compare_genes <- merge(rowData(sce_th_illumina), rowData(sce_th_mgi), by="ID", suffixes = c("_illumina", "_mgi"))
```


```{r venngenesTH}
venn.diagram(list(rownames(sce_th_mgi), rownames(sce_th_illumina)), 
             
             #aes
             col = colour_blue_yellow,
             fill = c(alpha(colour_blue_yellow[1]), alpha(colour_blue_yellow[2])),
             #save
             height = 4800,
             width = 7000,
             resolution = 300,
             filename = here(save_path, "vennTHilluminavsMGI_genes.png"),
             disable.logging=TRUE,
             #text nums
             cex=4,
             fontfamily="sans",
             #text categories
             category.names = c("MGI" , "Illumina"),
             cat.cex = 4,
             cat.fontfamily="sans",
             cat.default.pos="outer",
             cat.dist = c(0.055, 0.055),
             #title
             main="Genes",
             main.fontfamily = "sans",
             main.cex=5
             )
```

## correlation genes
```{r}
lm <- lm(compare_genes$mean_illumina ~ compare_genes$mean_mgi)
r2 <- summary(lm)$r.squared
r2

ggplot(as.data.frame(compare_genes), aes(x=mean_illumina, y=mean_mgi)) +
  geom_point(shape="o") +
  ggtitle("Gene correlation") +
    labs(x= "Illumina", y= "MGI",
       caption = "Average expression of each gene across all cells") +
 # geom_smooth(method = "lm", colour = "black")
  geom_text(x = 100, y = 250, label = expression(paste("R"^2, "= 0.9958"))) +#, round(r2,4)))) # I really tried to use the 
  theme_classic()
  
  
# same colouring outliers
as.data.frame(compare_genes) %>% 
  mutate(mean_ratio = mean_illumina / mean_mgi) %>% 
  mutate(outlier = ifelse(mean_ratio < quantile(mean_ratio, 0.025) | mean_ratio > quantile(mean_ratio, 0.975), "TRUE", "FALSE")) %>% 
  arrange(outlier) %>% # there are all condensed in low expression area, so we need to put them on top so we see them. 
ggplot(aes(x=mean_illumina, y=mean_mgi, colour = outlier )) +
  geom_point(shape="o") +
  ggtitle("Gene correlation", 
          subtitle = paste("Num Genes =", dim(compare_genes)[1])) +
    labs(x= "Illumina", y= "MGI",
       caption = "Average expression of each gene across all cells") +
 # geom_smooth(method = "lm", colour = "black")
 # geom_text(x = 100, y = 250, label = expression(paste("R"^2, "= 0.9958"))) +#, round(r2,4)))) # I really tried to use the 
  theme_classic()
```

Same plot without the genes with highest expression
```{r}
as.data.frame(compare_genes) %>% 
  filter(mean_mgi != max(mean_mgi))  %>% 
ggplot(aes(x=mean_illumina, y=mean_mgi )) +
   geom_point(shape="o") +
  ggtitle("Gene correlation", 
          subtitle = paste("Num Genes =", dim(compare_genes)[1]-1)) +
    labs(x= "Illumina", y= "MGI",
       caption = "Average expression of each gene across all cells - With no MALAT1") +
  geom_text(x = 30, y = 75, label = expression(paste("R"^2, "= 0.9958"))) +
    theme_classic()

```


The outliers are very low expression genes
```{r}
as.data.frame(compare_genes) %>% 
  mutate(mean_ratio = mean_illumina / mean_mgi) %>% 
  mutate(outlier = ifelse(mean_ratio < quantile(mean_ratio, 0.025) | mean_ratio > quantile(mean_ratio, 0.975), "outlier", ".")) %>% 
  filter(outlier == "outlier")  %>% 
ggplot(aes(x=mean_illumina, y=mean_mgi )) +
  geom_point() +
  ggtitle("Outlier genes")
```

